Preventing a Nuclear ISIL: \\Recommendations for United States Policy

### 2015-05-11 draft.tex
Added in subsection 6.6.4 - "Bilateral Nuclear Security Agreements."

% [3] “Iraq Weapons Inspections Fast Facts,” CNN, http://www.cnn.com/2013/10/30/world/meast/iraq-weapons-inspections-fast-facts/, 14 March, 2015.
% [4] Sidhu, W.P.S., “GCSP Policy Brief No. 19: Proliferation, Non-state Actors, and the Impact on Global Security,” Geneva Centre for Security Policy, 6 December, 2006.
% [5] Allison, G., Nuclear Terrorism: The Ultimate Preventable Catastrophe, New York: Henry Holt and Company, 2004.
% [6] “Battle for Iraq and Syria in maps,” BBC, \url{<http://www.bbc.com/news/world-middle-east-27838034>}, 24 April, 2015.
% [7] “World Map - Political,” \url{http://geology.com/world/world-map.shtml}, Accessed: 3 May, 2015.
% [8] Shatz, H.J., “How ISIS Funds Its Reign of Terror,” Rand Corp., New York Daily News, http://www.rand.org/blog/2014/09/how-isis-funds-its-reign-of-terror.html, 8 September, 2014.

@@

Russia

The U.S. signed the New Strategic Arms Reduction Treaty (New START) with Russia on 8 April 2010, and the treaty entered into force on 5 February 2011. The treaty limits the number of deployed strategic (SNW) to 1,550 on each side, to be achieved by February 2018. Deployed ICBMs, SLBMs, and heavy bombers assigned to nuclear missions are limited to 700, and deployed and non-deployed ICBM launchers, SLBM launchers, and bombers are limited to 800. Non-deployed ICBMs and SLBMs will also be monitored through on-site inspections and must be located at specified facilities away from deployment sites; labeled with unique identifiers [1].

New START provided a verification mechanism that allow both sides to keep check on each other's nuclear capabilities, building transparency and confidence. Importantly, the treaty will also keep account of all SNM/warheads that exists, not only for the benefit of US-Russia nuclear relations but also the possibility of nuclear weapons proliferation at the hands of non-state actors like ISIL.

Policy recommendations:

The US should maintain New START as one of the options to strengthen the overall verification process of nuclear weapons, missiles and launchers as means to ensure global nuclear weapons security. New Start has been under the purview of the US Department of State, working in conjunction with the US Department of Defense.

Up to 2012,  the cost of implementation of the New START has been \$301 million USD [2]. Although the New START is not put into effect with the aim of preventing ISIL, it will have have positive complementary effects. Presumably, it will make it more difficult for ISIL to obtain and transport nuclear weapons and material through theft because of increased nuclear security.

Pakistan

One of the most troublesome aspects of Pakistan's highly vulnerable nuclear security facilities is the high likelihood of theft by ISIL. Although the US has been very supportive of Pakistan's efforts to ensure its nuclear weapons security, there are various challenges that could still undermine Pakistan's efforts. Pakistan has received various military and non-military aid, close to 28.4 billion USD post 9/11 [3].

Policy recommendations:

The US need to continue to assist Pakistan in increasing its nuclear security capabilities, by providing monetary assistance and military hardware, as well as services aid like training and intelligence-sharing. The U.S. Department of State has been the lead agency in maintaining the US-Pakistan bilateral relations; however, when military specific assistance is needed, DoS will work closely with the U.S. Department of Defense on this issue.

North Korea

@@

The absence of diplomatic relations with North Korea stems from North Korea's attempted annexation of South Korea resulting in the Korean War and US imposed economic sanctions on North Korea [4]. The only communication between the two countries has been under the Six-Party Talks for the Denuclearization of the Korean Peninsula. Most forms of U.S. economic assistance, other than purely humanitarian assistance, are prohibited; therefore, U.S. assistance has been limited to food or energy aid. Between 1995 and 2001, the US has spent more than \$1.3 billion USD in total aid to North Korea (food aid - \$708 million USD , energy assistance - \$404 million USD, Six-Party Talks-related assistance - \$191 million USD) [5].

Policy recommendations:

The U.S. should continue the Six-Party Talks and begin actively engaging North Korea. Historical memory and isolation from the international world has made North Korea extremely hostile to the U.S. However, North Korea desires international legitimacy and has claimed that it wants to be a responsible nuclear power. Building upon that objective, North Korea and the U.S. can form multilateral agreements with China and South Korea aimed at preventing theft or sale of weapons through securing nuclear weapons facilities. Eventually, conditional agreements can be put in place where North Korea allows IAEA inspectors in exchange for food or energy.

Iran

The U.S. also need to address Iran's nuclear ambitions.  The U.S. is working under P5+1 initiatives to negotiate a nuclear deal to halt Iran's nuclear development. This cooperation will ensure monitoring and verification of Iran's nuclear development. Although it is known that Iran and ISIL have opposing ambitions and will likely not cooperate together, Iran's nuclear capabilities could also lead to proliferation of nuclear weapons into non-state actors and future proxy wars involving nuclear weapons.

Policy recommendations

@@

% References:
% [1] New Start, US Department of State (http://www.state.gov/t/avc/newstart/)
% [2] Report on Plan to Implement the Nuclear Force Reductions, Limitations, and Verification and Transparency Measures Contained in the New START Treaty Specified in Section 1042 of the National Defense Authorization Act for Fiscal Year 2012 (U)
% [3] Upadhyay, Brajesh, ‘US and Pakistan: Will the relationship endure?', BBC News, 19 Nov 2014 (http://www.bbc.com/news/world-us-canada-30105416)
% [4] US Relations with North Korea Fact Sheet (http://www.state.gov/r/pa/ei/bgn/2792.htm)
% [5]Manyin, Mark E. and Nikitin, Mary Beth, ‘Foreign Assistance to North Korea', Congressional Research Service, 2 April 2012
% [6] The Enemy of Our Enemy: Iran and ISIS on a Collision Course (http://www.breitbart.com/national-security/2014/09/17/the-enemy-of-our-enemy-iran-and-isis-on-a-collision-course/)

>>>>

@@

\textbf{Russia}

The U.S. signed the New Strategic Arms Reduction Treaty (New START) with Russia on 8 April 2010, and the treaty entered into force on 5 February 2011. The treaty limits the number of deployed strategic nuclear weapons to 1,550 on each side, to be achieved by February 2018. Deployed ICBMs, SLBMs, and heavy bombers assigned to nuclear missions are limited to 700, and deployed and non-deployed ICBM launchers, SLBM launchers, and bombers are limited to 800. Non-deployed ICBMs and SLBMs will also be monitored through on-site inspections and must be located at specified facilities away from deployment sites, labeled with unique identifiers \cite{U.S.DepartmentofState2015}.

New START provides a verification mechanism that allows both sides to keep check on each other's nuclear capabilities, building transparency and confidence. Importantly, the treaty will also keep account of all SNM/warheads that exist, not only for the benefit of U.S.-Russia nuclear relations but also for the possibility of nuclear weapons proliferation at the hands of non-state actors like ISIL.

\textbf{Policy recommendations:}

The U.S. should maintain New START as one of the options to strengthen the overall verification process of nuclear weapons, missiles and launchers, as a means to ensure global nuclear weapons security. New START has been under the purview of the U.S. Department of State, working in conjunction with the U.S. Department of Defense.

Up to 2012,  the cost of implementation of  New START has been U.S. \$301 million \cite{U.S.DepartmentofDefense2012}. Although  New START is not put into effect with the aim of preventing ISIL, it will have have positive complementary effects. Presumably, it will make it more difficult for ISIL to obtain and transport nuclear weapons and material through theft because of increased nuclear security.

\textbf{Pakistan}

One of the most troublesome aspects of Pakistan's highly vulnerable nuclear  facilities is the high likelihood of theft by ISIL. Although the U.S. has been very supportive of Pakistan's efforts to ensure its nuclear weapons security, there are various challenges that could still undermine Pakistan's efforts. Pakistan has received various military and non-military aid, close to U.S. \$28.4 billion  since 9/11 \cite{Upadhyay2014}.

\textbf{Policy recommendations:}

The U.S. needs to continue to assist Pakistan in increasing its nuclear security capabilities, by providing monetary assistance and military hardware, as well as services such as training and intelligence sharing. The U.S. Department of State has been the lead agency in maintaining the U.S.-Pakistan bilateral relations; however, when military-specific assistance is needed, DoS will work closely with the U.S. Department of Defense on this issue.

\textbf{North Korea}

@@

The absence of diplomatic relations with North Korea stems from North Korea's attempted annexation of South Korea, resulting in the Korean War and U.S. imposed economic sanctions on North Korea \cite{U.S.DepartmentofState2015a}. The only communication between the two countries has been under the Six-Party Talks for the Denuclearization of the Korean Peninsula. Most forms of U.S. economic assistance, other than purely humanitarian assistance, are prohibited; therefore, U.S. assistance has been limited to food or energy aid. Between 1995 and 2001, the U.S. has spent more than U.S. \$1.3 billion  in total aid to North Korea (food aid - U.S. \$708 million, energy assistance - U.S. \$404 million, Six-Party Talks-related assistance - U.S. \$191 million) \cite{Manyin2014}.

\textbf{Policy recommendations:}

The U.S. should continue the Six-Party Talks and begin actively engaging North Korea. Historical memory and isolation from the international world has made North Korea extremely hostile to the U.S. However, North Korea desires international legitimacy and has claimed that it wants to be a responsible nuclear power. Building upon that objective, North Korea and the U.S. can form multilateral agreements with China and South Korea aimed at preventing theft or sale of weapons through securing nuclear weapons facilities. Eventually, conditional agreements could be put in place where North Korea allows IAEA inspectors, in exchange for food or energy.

\textbf{Iran}

The U.S. also need to address Iran's nuclear ambitions.  The U.S. is working under P5+1 initiatives to negotiate a nuclear deal to halt Iran's nuclear development. This cooperation will ensure monitoring and verification of Iran's nuclear development. Although it is known that Iran and ISIL have opposing ambitions and will likely not cooperate together, Iran's nuclear capabilities could also lead to proliferation of nuclear weapons into non-state actors and future proxy wars involving nuclear weapons \cite{Pfarrer2014}.

\textbf{Policy recommendations:}

@@
